Kindly Bent to Free Us

### 2020-04-23 kindly/icfp20-rebuttal.md

We thank the reviewers for their thoughtful comments.
We will take their comments into account and revise our article as follows.
- Expand and improve our examples
- Precise further the limitations of our system and the comparison to other systems

# Reviewer A

> The use of "absolute" natural numbers to identify regions seems not modular.

Affe requires that each region is identified by an index along with a partial
order on indices that is compatible with the nesting of regions.
This can be implemented in many ways, including region variables
often used in algebraic effects systems, existentials, etc.

Natural numbers are the absolute simplest way to implement such indices
which is why
we use them in the formalisation. Our formalisation only relies on
the partial order and could easily use any of these
more sophisticated systems. This would certainly be the case in a 
realistic implementation for OCaml, for instance, to leverage
the on-going work on algebraic effects.
The properties of the system would stay identical.

We plan to add this point to section 6.

> It is very hard to get a grasp what kind of programming idioms can be type checked using Affe.

Affe can easily express rich data-structures (notably 
persistent ones) or manipulate resources and protocols (for instance with session types).

The main limitations of Affe is that it is not flow sensitive.
Code idioms that relies of subtle flow-sensitive usage of 
permissions and linearity will not work in Affe. Such patterns often
do not work in Rust either, and require richer logic more akin to theorem
proving, such as Mezzo.

Rust slightly relaxes these constraints through the use of non-lexical lifetimes
and double-borrows. 
Following the work by Weiss et al. [38] on the Rust semantics, we believe
these techniques can also be applied to Affe and 
sketch how in section 6.

> What is the reason the borrow is fixed to `U` in the splitting rules `SuspB` and `SuspS` (page 10).

Peter/Hannes ? Is this due to a proof thingy ? In principle, doing an exclusive reborrow is ok.

> Is Affe's kind mechanism also powerful enough to keep track of things like `Send` and `Sync`?

This is probably better implemented via ad-hoc polymorphism. It is very
desirable to extend Affe with some form of typeclasses, as we point out in
Section 6.2, which would allow to implement such mechanisms.

# Reviewer B

> What could go wrong, for example, if the set function of the array module would use a non-exclusive borrow type in the argument?

This question should be reframed in a larger context: "What does my API guarantee ?".
For instance, ML references and arrays allow concurrent writes in a perfectly
safe manner. In a more general concurrent context, 
many data-structures are designed to allow concurrent mutations, via mutexes for
instance.

However, if the API prevents concurrent mutations, this ensures that no data-races
can happen on this data-structure. Data-races are often the source of subtle
and hard-to-fix bugs.
Exclusive borrow allows the author to choose if they want or not to provide
such guarantee. This has been used with huge success in the Rust community to 
write safe and fast concurrent programs.

# Reviewer C

> in other systems, one might use existential quantification tricks or magic wands to express temporarily borrowing an *element* of an array before returning the element to the array. What would be the equivalent here?

Indeed, this is something that we would like to add to the system. 
Such functions can already be implemented via closures, 
as demonstrated in the iteration examples in section 2.3:

    val get_borrow :
      ('b <= affₖ) < &('a t) * int * (&(linₖ₊₁, 'a) -{linₖ₊₁}> 'b) -> 'b

This API is however a bit inconvenient. We can improve it using
so called "existential tricks" to allow matching on borrows:

    let &x = get_borrow (a,3) in (* &x must not escape here *)

In general, existentials tricks are not incompatible with Affe and are in fact
one of the foremost way to extend its expressivity.

> Do you envision a similar discipline for your system where one might write, say, a doubly-linked list in regular GC'd ML, then assume an interface for it that exposes it as a resource?

This pattern is indeed essential and was well demonstrated by Alms.
Affe already allow such patterns via subkinding and module type abstraction:
a type can be defined as unrestricted and exposed as linear.

Integrating "forced freeing" of such objects with the GC is more delicate
and is explored by Munch-Maccagnoni [21].

### 2020-04-25 kindly/icfp20-rebuttal.md

We thank the reviewers for their thoughtful comments.
We will take their comments into account and revise our article as follows.
- Expand and improve our examples
- Precise further the limitations of our system and the comparison to other systems

# Reviewer A

> The use of "absolute" natural numbers to identify regions seems not modular.

Affe requires that each region is identified by an index, along with a partial
order on indices that is compatible with the nesting of regions.
This can be implemented in many ways, including region variables
often used in algebraic effect systems, existentials, etc.

Natural numbers are the absolute simplest way to implement such indices
which is why
we use them in the formalization. Our formalization only relies on
the partial order and could easily use any of these
more sophisticated systems. This would certainly be the case in a 
realistic implementation for OCaml, for instance, to leverage
the ongoing work on algebraic effects.
The properties of the system would stay identical.

@@

We plan to add this point to section 6.

> It is very hard to get a grasp what kind of programming idioms can be type checked using Affe.

Affe can easily express rich data-structures (notably 
persistent ones) or manipulate resources and protocols (for instance with session types).

The main limitations of Affe is that it is not flow sensitive.
Code idioms that relies on subtle flow-sensitive usage of 
permissions and linearity will not work in Affe. Such patterns often
do not work in Rust either, and require richer logic more akin to theorem
proving, such as Mezzo.

@@

This is probably better implemented via ad-hoc polymorphism. It is very
desirable to extend Affe with some form of typeclasses, as we point out in
Section 6.2, which would allow to implement such mechanisms.

# Reviewer B

> What could go wrong, for example, if the set function of the array module would use a non-exclusive borrow type in the argument?

This refers to a larger question: "What does my API guarantee ?".
ML references and arrays allow concurrent writes but are subject to data-races.
Data-races are the source of subtle and hard-to-fix bugs.

One can prevent such issues dynamically with mutexes. Exclusive borrows
aims to forbid concurrent mutations *statically*. The choice then falls on the 
API designer: a file handle is probably better checked statically.
A distributed queue on the other hand would 
feature a `push` function taking a shared borrow precisely
to allow concurrent writes.

While a full description of a concurrent story for Affe is out of scope 
of this paper, such concepts have been used with huge success in the Rust
community to write safe and fast concurrent programs.

@@

# Reviewer C

> in other systems, one might use existential quantification tricks or magic wands to express temporarily borrowing an *element* of an array before returning the element to the array. What would be the equivalent here?

Indeed, this is something that we would like to add to the system. 
Such functions can already be implemented via closures, 
as demonstrated in the iteration examples in section 2.3:

    val get_borrow :
      ('b <= linₖ) => &('a t) * int * (&(linₖ₊₁, 'a) -{linₖ₊₁}> 'b) -> 'b

@@

This API is however a bit inconvenient. We can improve it using
so called "existential tricks" to allow matching on borrows:

    let &x = get_borrow (a,3) in (* &x must not escape here *)

In general, existential types are not incompatible with Affe and are in fact
one of the first way to extend its expressivity. They also happen to be already implemented in several ML-family languages (OCaml, notably).

@@

> Do you envision a similar discipline for your system where one might write, say, a doubly-linked list in regular GC'd ML, then assume an interface for it that exposes it as a resource?

This pattern is indeed essential and was well demonstrated by Alms.
Affe already allows such patterns via subkinding and module type abstraction:
a type can be defined as unrestricted and exposed as linear.

Integrating "forced freeing" of such objects with the GC is more delicate,
and we hope to reuse the work by Munch-Maccagnoni [21] for this purpose.

### 2020-04-25 kindly/icfp20-rebuttal.md

We thank the reviewers for their thoughtful comments.
We will take their comments into account and revise our article as follows.
- Expand and improve our examples
- Provide a comparison table with other systems 
- Further discuss limitations of our system

@@

# Reviewer A

> The use of "absolute" natural numbers to identify regions seems not modular.

Affe requires that each region is identified by an index drawn from a partial
order that is compatible with the nesting of regions.
This order can be implemented in many ways, including region variables
as often used in algebraic effects systems, existentials, etc.

For simplicity, the formalization uses the concrete implementation with 
natural numbers for indices. Our proofs only rely on
the partial order and could be adapted to one of the
more abstract approaches.

@@

We plan to add this point to section 6.

> It is very hard to get a grasp what kind of programming idioms can be type checked using Affe.

Affe can easily express rich data-structures (notably 
persistent ones) or manipulate resources and protocols (for instance with session types).

The main limitation of Affe is that it is not flow sensitive.
Code idioms that rely on subtle flow-sensitive usage of 
permissions and linearity do not typecheck in Affe. Such patterns often
do not work in Rust either and require a richer logic as in Mezzo.

@@

Rust slightly relaxes these constraints through the use of non-lexical lifetimes
and double-borrows. 
Following the work by Weiss et al. [38] on the Rust semantics, we believe
these techniques can also be applied to Affe and 
sketch how in section 6.

> What is the reason the borrow is fixed to `U` in the splitting rules `SuspB` and `SuspS` (page 10).

For SuspB it should be ok to relax the `U` to `b`. For SuspS, the idea
is that an exclusive borrow is affine. It would most likely be fine in
the present system, but it wouldn't scale to a concurrent setting.
We will check the proofs, implement the most general setting in the
rule, and comment on possible restrictions in a concurrent setting.

@@

This is probably better implemented via ad-hoc polymorphism. It is very
desirable to extend Affe with some form of typeclasses, as we point out in
Section 6.2, which would allow to implement such mechanisms.

# Reviewer B

> What could go wrong, for example, if the set function of the array module would use a non-exclusive borrow type in the argument?

This question should be reframed in a larger context: "What guarantee
is offered by an API?".
For instance, ML references and arrays allow concurrent writes in a perfectly
safe manner. In a more general concurrent context, 
many data-structures are designed to allow concurrent mutations, via mutexes for
instance.

@@

> in other systems, one might use existential quantification tricks or magic wands to express temporarily borrowing an *element* of an array before returning the element to the array. What would be the equivalent here?

Indeed, this is something that we would like to add to the system. 
Such functions can already be implemented via closures, 
as demonstrated in the iteration examples in section 2.3:

    val get_borrow :
      ('b <= affₖ) < &('a t) * int * (&(linₖ₊₁, 'a) -{linₖ₊₁}> 'b) -> 'b

This API is however a bit inconvenient. We can improve it using
so called "existential tricks" to match on borrows:

@@

    let &x = get_borrow (a,3) in (* &x must not escape here *)

In general, existentials tricks are not incompatible with Affe and are in fact
one of the foremost way to extend its expressivity.

> Do you envision a similar discipline for your system where one might write, say, a doubly-linked list in regular GC'd ML, then assume an interface for it that exposes it as a resource?

This pattern is indeed essential and is well demonstrated by Alms.
Affe allows such patterns via subkinding and module type abstraction:
a type can be defined as unrestricted and exposed as linear.

### 2020-04-25 kindly/icfp20-rebuttal.md

This is probably better implemented via ad-hoc polymorphism. It is very
desirable to extend Affe with some form of typeclasses, as we point out in
Section 6.2, which would allow to implement such mechanisms.

# Reviewer B

> What could go wrong, for example, if the set function of the array module would use a non-exclusive borrow type in the argument?

The real question here is: "What guarantee is offered by an API?".
ML references and arrays permit concurrent writes but are prone to
data races, which give rise to subtle bugs.

One can prevent such issues dynamically with mutexes, whereas exclusive borrows
forbid concurrent mutations *statically*. The choice then falls on the 
API designer: a file handle is probably better checked statically.
A distributed queue on the other hand would 
feature a `push` function taking a shared borrow precisely
to permit concurrent writes.

While a full description of a concurrent story for Affe is out of scope 
of this paper, such concepts have been used with huge success in the Rust
community to write safe and fast concurrent programs.

@@

    let &x = get_borrow (a,3) in (* &x must not escape here *)

In general, existential types are not incompatible with Affe and are in fact
one of the first way to extend its expressivity. They also happen to be already implemented in several ML-family languages (OCaml, notably).

> Do you envision a similar discipline for your system where one might write, say, a doubly-linked list in regular GC'd ML, then assume an interface for it that exposes it as a resource?

This pattern is indeed essential and is well demonstrated by Alms.
Affe supports such patterns via subkinding and module type abstraction:
a type can be defined as unrestricted and exposed as linear.

Integrating "forced freeing" of such objects with the GC is more delicate,
and we hope to reuse the work by Munch-Maccagnoni [21] for this purpose.

### 2020-04-25 kindly/icfp20-rebuttal.md

We plan to add this point to section 6.

> It is very hard to get a grasp what kind of programming idioms can be type checked using Affe.

Affe can easily express rich data-structures (notably 
persistent ones) or manipulate resources and protocols (for instance with session types).

The main limitation of Affe is that it is not flow sensitive.
Code idioms that rely on subtle flow-sensitive usage of 
permissions and linearity do not typecheck in Affe. Such patterns often
do not work in Rust either and require a richer logic as in Mezzo.
As a simple example, Affe can not directly type functions such as the 
merge on linear lists:
```
let rec merge l1 l2 = match l1, l2 with
  | h1::t1, h2::t2 ->
    if &h1 < &h2 
    then h1::(merge t1 l2) (* Must expand l2 to h2::t2 here *)
    else h2::(merge l1 t2)
  | ....
```

Rust slightly relaxes these constraints through the use of non-lexical lifetimes
and double-borrows. 
Following the work by Weiss et al. [38] on the Rust semantics, we believe
these techniques can also be applied to Affe and 
sketch how in section 6, which would help for the example above.

### 2020-05-15 kindly/intro.tex
first round of revision

Direct style means that there is a function like @fopen@ that creates
and returns a linear resource. In contrast, LH forces programmers to
use resource-passing style because, in LH, linearity is a property of
a function, rather than a property of a value that restricts the way
that value can be handled (as in \lang). An LH API analogous to @File@
might provide functions like
\begin{itemize}
\item @withFile : path -> (handle o-> r) -> r@, which creates a new
  file handle and takes a continuation that uses the @handle@ linearly,
\item @writeFile : string -> handle o-> handle@, which
  returns the transformed resource @handle@, and
\item @closeFile : handle o-> unit@, which consumes the @handle@ by
  closing the file.
\end{itemize}

@@

\begin{itemize}
\item A type system encoding linearity and affinity with
  borrowing. The type system is polymorphic for types and
  for kinds that express constraints on the number of uses of a value.
\item An extension of the \hmx framework
  \cite{DBLP:journals/tapos/OderskySW99} with kinds, regions, and
  multiplicity constraints that supports full, principal type inference.
\item Soundness proof of the inference algorithm.
\item Big-step linearity aware semantics with an expressive soundness theorem.
\item Constraint simplification techniques.
\item Automatic inference of regions for borrows.
\item A prototype implementation of the type inference algorithm, including all
  constraint simplification and extended with algebraic datatypes and
  pattern matching,
  available at \url{https://affe.netlify.com/}\footnote{This link is safe for double-blind.}.
\end{itemize}

>>>>

Direct style means that there is a function like @fopen@ that creates
and returns a linear resource. In contrast, LH forces programmers to
use resource-passing style because, in LH, linearity is a property of
a function, rather than a property of a value that restricts the way
that value can be handled (as in \lang). An LH API analogous to @File@
might provide functions like
\begin{itemize}
\item @withFile : path -> (handle o-> Unrestricted r) -> r@, which creates a new
  file handle and takes a continuation that uses the @handle@
  linearly, but returns an unrestricted value,
\item @writeFile : string -> handle o-> handle@, which
  returns the transformed resource @handle@, and
\item @closeFile : handle o-> unit@, which consumes the @handle@ by
  closing the file.
\end{itemize}

@@

\begin{itemize}
\item A polymorphic type system that encodes linearity and affinity with
  borrowing in lexical regions. Polymorphism covers types and kinds that
  express multiplicity constraints on the number of uses of a
  value. This type system is a conservative extension of systems for
  existing ML-like languages.
\item Expressive type soundness theorem with respect to a big-step linearity-aware semantics. 
\item An extension of the \hmx framework
  \cite{DBLP:journals/tapos/OderskySW99} for constrained type
  inference to equip the type system with full, principal type inference.
\item Soundness proof of the inference algorithm.
% \item Constraint simplification techniques.
\item Automatic inference of regions for borrows.
\item A prototype implementation of the type inference algorithm, including all
  constraint simplification and extended with algebraic datatypes and
  pattern matching,
  available at \url{https://affe.netlify.com/}.
\end{itemize}

### 2020-05-15 kindly/lang.tex
first round of revision

A borrow $\borrow{x}$ is always taken from a variable $x$. The
\emph{borrow annotation} $\BORROW$ indicates whether the borrow is exclusive
($\MBORROW$) or shared ($\IBORROW$). If $x$ is already borrowed,
then we need to use the reborrow expression $\reborrow{x}$.
%
A region $\region{\Sone x\BORROW}{e}$ is annotated with its nesting $n$, the variable $x$ that may be borrowed in $e$, and the kind of borrow $\BORROW$.
%
A match is indexed with a \emph{match specification} $\etransfm$ that indicates
whether the match operates on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
%
We consider four primitive operations to manipulate resources:
$\create$, $\observe$, $\update$ and $\destroy$.

>>>>

A borrow $\borrow{x}$ is always taken from a variable $x$. The
\emph{borrow annotation} $\BORROW$ indicates whether the borrow is exclusive/affine
($\MBORROW$) or shared/unrestricted ($\IBORROW$). If $x$ is already borrowed,
then we need to use the reborrow expression $\reborrow{x}$.
%
A region $\region{\Sone x\BORROW}{e}$ is annotated with its nesting $n$, the variable $x$ that may be borrowed in $e$, and the kind of borrow $\BORROW$.
%
A match is indexed with a \emph{match specification} $\etransfm$ that indicates
whether the match operates on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
%
We consider four primitive operations to manipulate resources:
$\create$, $\observe$, $\update$ and $\destroy$.
They serve a prototypes to demonstrate the typing and handling of
resources. For a concrete type of resource, there are further
arguments and perhaps different versions of the operations. But the
typing and behavior of the operations is analogous to the prototype
operations.

### 2020-05-15 kindly/motivation-persistent.tex
first round of revision

\cref{sig:hybarray} contains an API @HYBARRAY@
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the @Array@
signature. For simplicity, the array elements are unrestricted, but
the structure itself remains linear so as to be implemented in terms of @Array@.
The in-place mutation function @set_mut@ with type
@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
@set@ operation has type @&('a t) * int * 'a -> 'a t@. It
takes a shared borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
@HYBARRAY@ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
@mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
is a simple variation on @Array.map@ where the mapping function also
takes the position of the element.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

\begin{figure}[tp]
  \centering
  \begin{subfigure}[t]{0.45\linewidth}
\begin{lstlisting}
module type HYBARRAY = sig
  include ARRAY
  val set : &('a t) * int * 'a -> 'a t
  val set_mut : &!('a t) * int * 'a -> unit
end
\end{lstlisting}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.55\linewidth}
\begin{lstlisting}
module Cow : HYBARRAY = struct
  include Array
  let set (a, i0, x0) =
    Array.mapi (fun i x -> if i = i0 then x0 else x) a
  let set_mut = Array.set
end
\end{lstlisting}
  \end{subfigure}
  \vspace{-15pt}
  \caption{Signature and Implementation of hybrid arrays}
  \label{sig:hybarray}
  \label{ex:cow}

  \begin{subfigure}[t]{0.45\linewidth}
  \lstinputlisting[linerange=1-13]{code/sukodu.affe}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.55\linewidth}
  \lstinputlisting[linerange=14-31]{code/sukodu.affe}
  \end{subfigure}
  % \lstinputlisting[linerange=1-31,multicols=2]{code/sukodu.affe}
  \vspace{-15pt}
  \caption{Excerpt of the Sudoku solver}
  \label{ex:sudoku}
\end{figure}

@@

The board is represented as a 2D-matrix (\cref{line:boardtype}), where
the @Matrix@ type uses the same API as @PersistArray@
but with two indices.
Each cell contains an integer set that represents admissible solutions so far.
The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we create a choice point,
change the cell with an immutable modification (\cref{line:immutset}), and propagate
the changes with the @propagate@ function.
The @propagate@ function uses direct mutation through an
exclusive borrow of the matrix as it need not preserve the previous
version of the board.
The implementation of @propagate@ is split into three parts
for lines, columns, and square, which are all very similar to function
@propagate_lines@ (\cref{line:propline}).

>>>>

\cref{sig:hybarray} contains an API @HYBARRAY@
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the @Array@
signature. As our application requires the @get@ function, the array
elements must be unrestricted, but
the structure itself remains linear so as to be implemented in terms of @Array@.
The in-place mutation function @set_mut@ with type
@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
@set@ operation has type @&('a t) * int * 'a -> 'a t@. It
takes a shared borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
@HYBARRAY@ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
@mapi: (int * &'a -> 'b) * &('a t) -> 'b t@
is a simple variation on @Array.map@ where the mapping function also
takes the position of the element.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

\begin{figure}[tp]
  \centering
  \begin{subfigure}[t]{0.45\linewidth}
\begin{lstlisting}
module type HYBARRAY = sig
  include ARRAY
  val set : &('a t) * int * 'a -> 'a t
  val set_mut : &!('a t) * int * 'a -> unit
end
\end{lstlisting}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.55\linewidth}
\begin{lstlisting}
module CowArray : HYBARRAY = struct
  include Array
  let set (a, i0, x0) =
    Array.mapi ((fun (i, x) -> if i = i0 then x0 else x), a)
  let set_mut = Array.set
end
\end{lstlisting}
  \end{subfigure}
  \vspace{-15pt}
  \caption{Signature and Implementation of hybrid arrays}
  \label{sig:hybarray}
  \label{ex:cow}

  \begin{subfigure}[t]{0.45\linewidth}
  \lstinputlisting[linerange=1-13]{code/sukodu.affe}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.55\linewidth}
  \lstinputlisting[linerange=14-31,firstnumber=last]{code/sukodu.affe}
  \end{subfigure}
  % \lstinputlisting[linerange=1-31,multicols=2]{code/sukodu.affe}
  \vspace{-15pt}
  \caption{Excerpt of the Sudoku solver}
  \label{ex:sudoku}
\end{figure}

@@

The board is represented as a 2D-matrix (\cref{line:boardtype}), where
the @Matrix@ type uses the same API as @CowArray@
but with two indices.
Each cell contains an integer set that represents admissible solutions so far.
The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we create a choice point,
change the cell with an immutable modification (\cref{line:immutset}), and propagate
the changes with the @propagate@ function.
The @propagate@ function uses direct mutation through an
exclusive borrow of the matrix as it need not preserve the previous
version of the board.
The implementation of @propagate@ is split into three parts
for lines, columns, and square, which are all very similar to function
@propagate_lines@ (\cref{line:propline}).

### 2020-05-15 kindly/motivation.tex
first round of revision

\begin{figure}[tp]
  \centering
\begin{lstlisting}
module Array : sig
  type ('a : 'k) t : lin
  val create : ('a : un) => int -> 'a -> 'a t
  val free : ('a : aff) => 'a t -> unit
  val length : &('a t) -> int
  val get : ('a : un) => &('a t) * int -> 'a
  val set : ('a : aff) => &!('a t) * int * 'a -> unit
  val map : (&'a -> 'b) -> &('a t) -> 'b t
end
\end{lstlisting}
  \vspace{-15pt}
  \caption{Linear arrays}
  \label{ex:array}
  \label{sig:array}
  \vspace{-10pt}
\end{figure}

The API for mutable linear arrays (\cref{sig:array})
aims to safely handle manual allocation and
deallocation of arrays that may contain affine elements.
A program would first use @create n v@ to create
an array of size @n@ initialized with value
@v@. The value @v@ must be unrestricted as it is duplicated to
initialize all array elements. Using the @map@ function we
can transform the unrestricted elements into linear (affine)
ones. The resulting array can be processed further using @map@ (or @set@ if
affine). The @get@ function is only applicable if the element type
is unrestricted as one element is duplicated. The @length@
function is always applicable. To @free@ an array the elements must be
affine.

@@

Using the API 
we can create an
array of Fibonacci numbers in an imperative coding style:
\begin{lstlisting}
let mk_fib_array n =
  let a = create n 1 in
  for i = 2 to n - 1 do
    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
    set (&!a, i, x)(*@\label{ex:array:set}*)
  done;
  a(*@\label{ex:array:return}*)
# mk_fib_array : int -> int Array.t
\end{lstlisting}

After
creation of the array, the presence of a borrow in the for loop
prevents access to the ``raw'' resource inside the loop's
body. \cref{ex:array:get} contains two shared borrows
in the same expression which forms a region by itself. These borrows
are split off the exclusive borrow used in \cref{ex:array:set} which
belongs to the next enclosing region corresponding to the loop body.
The whole array can be returned in \cref{ex:array:return} because  the
borrows are no longer in scope.
%
More precisely, here is is an annotated excerpt with regions explicitly
marked by braces @{| ... |}@:
\begin{lstlisting}[firstnumber=3]
  for i = 2 to n - 1 do {|
    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
    set &!a i x
  |} done;
\end{lstlisting}

@@

% We have seen that borrows are limited to a region.
In the examples so far, regions do not appear in type signatures.
But for certain programming idioms, we want to extend the scope of a
region across a function boundary.
For instance, how should we fold on an array of linear objects?
A naive fold function would only work on unrestricted content:
\begin{lstlisting}
val fold : ('a : un) => ('a -> 'b -> 'b) -> 'a array -> 'b -> 'b
\end{lstlisting}

If we want to work on linear and affine resources,
we must make sure those resources are not leaked in the result.
content. We obtain the following signature:
\begin{lstlisting}
val fold :
  ('b:'k),('k <= aff_r) =>
  (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a array) -> 'b -{'k_1}> 'b
\end{lstlisting}

@@

This approach is not sufficient if we want to mutate the array during iteration.
To this end, we need to take an exclusive borrow of the structure to iterate
on:
\begin{lstlisting}
val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
\end{lstlisting}

@@

While the distinction between mutable and immutable iteration functions
seems unfortunate, it is typical of
programming with borrows and is also present in the Rust standard library.
It enables the programmer to explicitly state how different iterations
may be composed and optimized.
\lang's region variables ensure that the content iterated on can never
be leaked outside of the iteration function.
This pattern is essential in many use cases of linearity such
as pools of linear objects (see \cref{tuto:pool}).

>>>>

\begin{figure}[tp]
  \centering
\begin{lstlisting}
module Array : sig
  type ('a : 'k) t : lin
  val create : ('a : un) => int * 'a -> 'a t
  val free : ('a : aff) => 'a t -> unit
  val length : &('a t) -> int
  val get : ('a : un) => &('a t) * int -> 'a
  val set : ('a : aff) => &!('a t) * int * 'a -> unit
  val map : (&'a -> 'b) * &('a t) -> 'b t
end
\end{lstlisting}
  \vspace{-15pt}
  \caption{Linear arrays}
  \label{ex:array}
  \label{sig:array}
  \vspace{-10pt}
\end{figure}

The API for mutable linear arrays (\cref{sig:array})
aims to safely handle manual allocation and
deallocation of arrays that may contain affine elements.
A program would first use @create (n, v)@ to create
an array of size @n@ initialized with value
@v@. The value @v@ must be unrestricted as it is duplicated to
initialize all array elements. Using the @map@ function we
can transform the unrestricted elements into linear (affine)
ones. The resulting array can be processed further using @map@ (or @set@ if
affine). The @get@ function is only applicable if the element type
is unrestricted as one element is duplicated. The @length@
function is always applicable. To @free@ an array the elements must be
affine.

@@

Using the API 
we can create an
array of Fibonacci numbers in an imperative coding style:
\begin{lstlisting}
let mk_fib_array n =
  let a = create (n, 1) in
  for i = 2 to n - 1 do
    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
    set (&!a, i, x)(*@\label{ex:array:set}*)
  done;
  a(*@\label{ex:array:return}*)
# mk_fib_array : int -> int Array.t
\end{lstlisting}

After
creation of the array, the presence of a borrow in the for loop
prevents access to the ``raw'' resource inside the loop's
body. In particular, the resource cannot be freed through a
borrow. \cref{ex:array:get} contains two shared borrows 
in the same expression which forms a region by itself. These borrows
are split off the exclusive borrow used in \cref{ex:array:set} which
belongs to the next enclosing region corresponding to the loop body.
The whole array can be returned in \cref{ex:array:return} because  the
borrows are no longer in scope.
%
More precisely, here is is an annotated excerpt with regions explicitly
marked by braces @{| ... |}@:
\begin{lstlisting}[firstnumber=3]
  for i = 2 to n - 1 do {|
    let x = {| get (&a, i-1) + get (&a, i-2) |} in(*@\label{ex:array:region:get}*)
    set (&!a, i, x)
  |} done;
\end{lstlisting}

@@

% We have seen that borrows are limited to a region.
In the examples so far, regions do not appear in type signatures.
But for certain programming idioms, we want to extend the scope of a
region across a function boundary.
For instance, how should we fold on an array of linear objects?
A naive fold function would only work on unrestricted content:
\begin{lstlisting}
val fold : ('a : un) => ('a -> 'b -> 'b) -> 'a Array.t -> 'b -> 'b
\end{lstlisting}

If we want to work on linear and affine resources,
we must make sure those resources are not leaked in the result.
content. We obtain the following signature:
\begin{lstlisting}
val fold :
  ('b:'k),('k <= aff_r) =>
  (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a Array.t) -> 'b -{'k_1}> 'b
\end{lstlisting}

@@

This approach is not sufficient if we want to mutate the array during iteration.
To this end, we need to take an exclusive borrow of the structure to iterate
on:
\begin{lstlisting}
val iter_mut : &!('a Array.t) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
\end{lstlisting}

@@

While the distinction between mutable and immutable iteration functions
seems unfortunate, it is typical of
programming with borrows and is also present in the Rust standard library.
It enables the programmer to explicitly state how different iterations
may be composed and optimized.
It also enables different implementations such as
using parallel iterations in the immutable case.
\lang's region variables ensure that the content iterated on can never
be leaked outside of the iteration function.
This pattern is essential in many use cases of linearity such
as pools of linear objects (see \cref{tuto:pool}).
